\section{Conclusion}

\subsection{Summary}
This project presents a comprehensive study of Transformer-based Named Entity Recognition (NER) for IT recruitment job descriptions. We construct a domain-specific dataset of 11{,}001 synthetic samples with token-level IOB annotations for five practical entity types: \textbf{ROLE}, \textbf{SKILL}, \textbf{LOC}, \textbf{EXP}, and \textbf{SALARY}. To ensure training quality, we apply a two-stage preprocessing pipeline including generation validation and tokenizer-aware label alignment for subword tokenization.

We evaluate three pre-trained Transformer backbones---\textbf{BERT-base}, \textbf{RoBERTa-base}, and \textbf{DistilBERT-base}---under six optimization settings: Base CE, Focal Loss, Dice Loss, CE + Dice, CRF, and CRF + Focal Loss. Across all architectures, the experimental results consistently show that \textbf{CE + Dice Loss} is the most effective objective. The overall best model is \textbf{RoBERTa-base + CE + Dice}, achieving an F1-score of \textbf{0.9223} with strong precision-recall balance and high token-level accuracy. BERT-base + CE + Dice follows closely (F1 = 0.9202), while DistilBERT-base + CE + Dice remains highly competitive (F1 = 0.9109) with substantially fewer parameters.

These findings demonstrate that hybrid loss design can outperform standalone objectives, and that model selection should consider both predictive quality and deployment constraints (latency, memory, and compute budget).

\subsection{Limitations}
Despite strong performance, this study has several limitations.

First, the dataset is \textbf{synthetic} and generated by an LLM. Although diversity constraints are imposed, generated text may not fully reflect the linguistic noise, ambiguity, and formatting irregularities found in real-world job postings. This can limit external validity.

Second, experiments are conducted on a \textbf{single domain} (IT recruitment) with a fixed entity schema. The conclusions may not transfer directly to other recruitment sectors (e.g., healthcare, finance) or broader information extraction tasks.

Third, we report results from a fixed train/test split and limited hyperparameter settings. While this supports fair model comparison, it does not fully capture variance across random seeds, data splits, or broader hyperparameter search spaces.

Fourth, the work primarily emphasizes model-level metrics (F1, precision, recall, accuracy) and does not include deeper \textbf{error taxonomy} analyses such as boundary errors, type confusion matrices, or robustness under noisy/abbreviated inputs.

Finally, the current study focuses on offline evaluation and does not include production-oriented measurements such as end-to-end latency under realistic traffic, drift monitoring, or human-in-the-loop correction workflows.

\subsection{Future Work}
Several promising directions can extend this work.

\begin{itemize}
    \item \textbf{Real-world data integration:} Combine synthetic data with manually annotated real job descriptions to improve robustness and reduce synthetic-to-real domain gap.
    \item \textbf{Cross-domain generalization:} Evaluate transfer performance to non-IT recruitment domains and multilingual settings.
    \item \textbf{Advanced modeling strategies:} Explore modern encoder families (e.g., DeBERTa variants), parameter-efficient fine-tuning (LoRA/adapters), and span-based or prompt-based NER formulations.
    \item \textbf{Improved imbalance handling:} Investigate alternative reweighting and calibration strategies (e.g., class-balanced loss, label smoothing, threshold tuning) given the consistent weakness of Focal Loss in this setting.
    \item \textbf{Data quality and active learning:} Introduce uncertainty-driven sampling and iterative annotation to prioritize informative examples and reduce labeling cost.
    \item \textbf{Deployment and MLOps evaluation:} Benchmark inference speed, memory usage, and throughput across hardware targets; design monitoring pipelines for data drift and model degradation.
    \item \textbf{Explainability and trust:} Add token-level attribution and error explanation tools to support recruiter-facing transparency and easier model debugging.
\end{itemize}

In conclusion, this work confirms that Transformer-based NER is highly effective for extracting structured recruitment information from unstructured job descriptions. Among the evaluated strategies, \textbf{RoBERTa-base with CE + Dice Loss} offers the best raw performance, while \textbf{DistilBERT with CE + Dice Loss} provides an attractive efficiency-performance trade-off for practical deployment.