\section{Experiments \& Results}

\subsection{Experimental Setup}

\paragraph{Dataset and Preprocessing.}
We use a custom NER dataset stored in JSONL format with token-level and tag-level annotations.
The dataset contains 11,001 training examples, which we split into training and test sets with an 80/20 ratio
(8,800 training, 2,201 test) using a fixed random seed of 42 for reproducibility.

Each example consists of a sequence of tokens and corresponding NER tags. During preprocessing, we encode
all unique tags into numeric IDs and compute class weights based on tag frequency in the training set to
address label imbalance. These weights are used in Focal Loss variants.

Tokenization is performed using model-specific \texttt{AutoTokenizer} with the \texttt{is\_split\_into\_words=True}
flag to handle word-to-subword alignment. For multi-subword tokens, we assign the label to the first subword
and mask subsequent subwords with a special ignore index (\texttt{-100}) to exclude them from loss and metric computation.

\paragraph{Training Hyperparameters.}
We train all models with consistent hyperparameters across architectures:
\begin{itemize}
  \item \textbf{Learning rate:} $2 \times 10^{-5}$ (standard for fine-tuning pre-trained Transformers)
  \item \textbf{Batch size:} 16 (training and evaluation)
  \item \textbf{Number of epochs:} 5
  \item \textbf{Weight decay:} 0.01 (L2 regularization)
  \item \textbf{Evaluation strategy:} per-epoch
  \item \textbf{Optimizer:} AdamW (default in Hugging Face Trainer)
\end{itemize}

\paragraph{Loss Functions and Training Objectives.}
We compare six training objectives:

\begin{enumerate}
  \item \textbf{Base\_CE (Baseline):} Standard cross-entropy loss applied to token classification logits.
  
  \item \textbf{Focal Loss:} Weighted focal loss to handle class imbalance:
  \begin{equation}
    \text{FL}(p_t) = -\alpha_t (1 - p_t)^\gamma \log(p_t)
  \end{equation}
  where $p_t$ is the model probability for the correct class, $\gamma = 2.0$ is the focusing parameter, and $\alpha_t$ is the class weight computed from training tag frequencies.
  
  \item \textbf{Dice Loss:} A soft intersection-over-union loss commonly used in segmentation:
  \begin{equation}
    \text{DiceLoss} = 1 - \frac{2 \sum_c p_c y_c + \epsilon}{\sum_c p_c + y_c + \epsilon}
  \end{equation}
  where $p_c$ and $y_c$ are softmax probabilities and one-hot targets per class, and $\epsilon = 10^{-6}$ is a smoothing constant.
  
  \item \textbf{CE + Dice Loss:} A combined loss summing cross-entropy and Dice loss to leverage both.
  
  \item \textbf{CRF:} A Conditional Random Field layer replacing softmax decoding to model label transitions explicitly. The CRF learns transition scores between tag pairs and finds the maximum-likelihood sequence during inference.
  
  \item \textbf{CRF + Focal Loss:} CRF decoding with sequence-level Focal Loss applied to the CRF negative log-likelihood.
\end{enumerate}

\paragraph{Evaluation Metrics.}
We evaluate all models using the \texttt{seqeval} metric, which computes entity-level metrics:
\begin{itemize}
  \item \textbf{Precision:} proportion of predicted entities that are correct.
  \item \textbf{Recall:} proportion of gold entities that are predicted correctly.
  \item \textbf{F1-Score:} harmonic mean of precision and recall.
  \item \textbf{Accuracy:} token-level accuracy (including padding/masked tokens).
\end{itemize}
Metrics are computed on the test set after each epoch, and we report the final epoch results.

\subsection{BERT Results}

\subsubsection{Model Configuration.}
We train BERT (bert-base-cased) for NER token classification. BERT-base is a robust bidirectional Transformer model with 12 layers, 768 hidden dimensions, and 110M parameters. The tokenizer is configured to preserve exact word boundaries and casing, which is critical for identifying proper nouns in entity extraction.

\subsubsection{Baseline: CE Loss.}

\begin{table}[htbp]
    \centering
    \begin{tabular}{l c c c c c}
        \hline
        \textbf{Epoch} & \textbf{Train Loss} & \textbf{Val Loss} & \textbf{F1} & \textbf{Precision / Recall} & \textbf{Acc.} \\
        \hline
        1 & 0.2089 & 0.0888 & 0.9021 & 0.8936 / 0.9108 & 0.9721 \\
        2 & 0.0697 & 0.0794 & 0.9140 & 0.9129 / 0.9151 & 0.9743 \\
        3 & 0.0528 & 0.0791 & 0.9136 & 0.9079 / 0.9193 & 0.9740 \\
        4 & 0.0432 & 0.0828 & 0.9163 & 0.9119 / 0.9208 & 0.9748 \\
        5 & 0.0333 & 0.0872 & 0.9172 & 0.9140 / 0.9204 & 0.9755 \\
        \hline
    \end{tabular}
    \caption{BERT with Base Cross-Entropy Loss (5 epochs).}
    \label{tab:bert_base_ce}
\end{table}

\noindent The baseline achieves a strong final F1-score of 0.9172. Validation loss reaches a minimum at epoch 3 (0.0791), yet the F1 metric continues to improve, demonstrating robust generalization despite mild token-level overfitting. Precision and recall remain excellently balanced (0.9140 vs. 0.9204), providing a highly reliable foundation.

\subsubsection{Focal Loss.}

\begin{table}[htbp]
    \centering
    \begin{tabular}{l c c c c c}
        \hline
        \textbf{Epoch} & \textbf{Train Loss} & \textbf{Val Loss} & \textbf{F1} & \textbf{Precision / Recall} & \textbf{Acc.} \\
        \hline
        1 & 0.1905 & 0.0420 & 0.7789 & 0.7155 / 0.8547 & 0.9303 \\
        2 & 0.0320 & 0.0366 & 0.8083 & 0.7568 / 0.8673 & 0.9466 \\
        3 & 0.0220 & 0.0379 & 0.8343 & 0.7900 / 0.8840 & 0.9507 \\
        4 & 0.0156 & 0.0361 & 0.8382 & 0.7961 / 0.8849 & 0.9548 \\
        5 & 0.0119 & 0.0384 & 0.8502 & 0.8120 / 0.8921 & 0.9589 \\
        \hline
    \end{tabular}
    \caption{BERT with Focal Loss ($\gamma = 2.0$, weighted by class frequency).}
    \label{tab:bert_focal}
\end{table}

\noindent Focal Loss yields the lowest F1 (0.8502), dropping 0.0670 points below the baseline. Despite achieving the lowest absolute validation loss (0.0384), it suffers severe precision degradation (0.8120). This paradox suggests that aggressively penalizing easy background tokens disrupts the model's discriminative threshold, leading to massive false positive predictions.

\subsubsection{Dice Loss.}

\begin{table}[htbp]
    \centering
    \begin{tabular}{l c c c c c}
        \hline
        \textbf{Epoch} & \textbf{Train Loss} & \textbf{Val Loss} & \textbf{F1} & \textbf{Precision / Recall} & \textbf{Acc.} \\
        \hline
        1 & 0.2571 & 0.1876 & 0.8792 & 0.8868 / 0.8716 & 0.9630 \\
        2 & 0.1822 & 0.1847 & 0.8870 & 0.8936 / 0.8805 & 0.9656 \\
        3 & 0.1754 & 0.1823 & 0.8916 & 0.8909 / 0.8924 & 0.9664 \\
        4 & 0.1725 & 0.1803 & 0.8938 & 0.8960 / 0.8916 & 0.9673 \\
        5 & 0.1680 & 0.1778 & 0.8959 & 0.8983 / 0.8934 & 0.9685 \\
        \hline
    \end{tabular}
    \caption{BERT with Dice Loss ($\epsilon = 10^{-6}$).}
    \label{tab:bert_dice}
\end{table}

\noindent Standalone Dice Loss achieves a competitive F1 of 0.8959 with highly stable, monotonically decreasing validation loss (0.1778). By directly optimizing structural overlap, it produces an exceptional precision-recall balance (P = 0.8983, R = 0.8934). However, it slightly underperforms the CE baseline, indicating a need for token-level gradient stability during optimization.

\subsubsection{CE + Dice Loss.}

\begin{table}[htbp]
    \centering
    \begin{tabular}{l c c c c c}
        \hline
        \textbf{Epoch} & \textbf{Train Loss} & \textbf{Val Loss} & \textbf{F1} & \textbf{Precision / Recall} & \textbf{Acc.} \\
        \hline
        1 & 0.4638 & 0.2608 & 0.9071 & 0.8973 / 0.9172 & 0.9728 \\
        2 & 0.2323 & 0.2412 & 0.9156 & 0.9128 / 0.9185 & 0.9747 \\
        3 & 0.2011 & 0.2396 & 0.9153 & 0.9111 / 0.9195 & 0.9753 \\
        4 & 0.1804 & 0.2427 & 0.9156 & 0.9096 / 0.9217 & 0.9749 \\
        5 & 0.1664 & 0.2453 & 0.9202 & 0.9182 / 0.9221 & 0.9759 \\
        \hline
    \end{tabular}
    \caption{BERT with Combined CE + Dice Loss.}
    \label{tab:bert_ce_dice}
\end{table}

\noindent The hybrid CE + Dice approach achieves the highest overall F1-score of 0.9202. This variant perfectly synergizes the optimization stability of cross-entropy with the boundary-awareness of Dice loss. It converges rapidly, maintains an outstanding precision-recall balance (0.9182 vs. 0.9221), and represents the most robust architecture evaluated.

\subsubsection{CRF Layer.}

\begin{table}[htbp]
    \centering
    \begin{tabular}{l c c c c c}
        \hline
        \textbf{Epoch} & \textbf{Train Loss} & \textbf{Val Loss} & \textbf{F1} & \textbf{Precision / Recall} & \textbf{Acc.} \\
        \hline
        1 & 5.8880 & 2.0161 & 0.9005 & 0.8946 / 0.9065 & 0.9716 \\
        2 & 1.6642 & 1.8119 & 0.9099 & 0.9068 / 0.9131 & 0.9738 \\
        3 & 1.2439 & 1.7899 & 0.9110 & 0.9055 / 0.9166 & 0.9736 \\
        4 & 1.0700 & 1.9179 & 0.9122 & 0.9042 / 0.9204 & 0.9741 \\
        5 & 0.8204 & 1.9225 & 0.9137 & 0.9081 / 0.9193 & 0.9747 \\
        \hline
    \end{tabular}
    \caption{BERT + CRF (no focal loss).}
    \label{tab:bert_crf}
\end{table}

\noindent The CRF variant converges smoothly to an F1 of 0.9137, trailing the CE baseline by a marginal 0.0035 points. While explicitly enforcing sequence-level transition validity, the lack of significant empirical gain suggests that BERT's deep contextual embeddings naturally predict highly valid sequences on this dataset without requiring restrictive transition matrices.

\subsubsection{CRF + Focal Loss.}

\begin{table}[htbp]
    \centering
    \begin{tabular}{l c c c c c}
        \hline
        \textbf{Epoch} & \textbf{Train Loss} & \textbf{Val Loss} & \textbf{F1} & \textbf{Precision / Recall} & \textbf{Acc.} \\
        \hline
        1 & 5.6647 & 1.7833 & 0.8983 & 0.8926 / 0.9041 & 0.9715 \\
        2 & 1.4498 & 1.5809 & 0.9092 & 0.9059 / 0.9125 & 0.9737 \\
        3 & 1.0559 & 1.5488 & 0.9110 & 0.9054 / 0.9167 & 0.9741 \\
        4 & 0.8976 & 1.6165 & 0.9099 & 0.9010 / 0.9191 & 0.9737 \\
        5 & 0.6526 & 1.6137 & 0.9118 & 0.9067 / 0.9169 & 0.9745 \\
        \hline
    \end{tabular}
    \caption{BERT + CRF with Focal Loss ($\gamma = 2.0$).}
    \label{tab:bert_crf_focal}
\end{table}

\noindent CRF + Focal Loss (0.9118 F1) underperforms both pure CRF and the baseline. This occurs because Focal Loss suppresses the confident token predictions that the CRF layer relies upon to establish robust structural transition paths, creating conflicting optimization signals that hinder overall sequence extraction.

\subsubsection{Summary: All BERT Variants.}

\begin{table}[htbp]
    \centering
    \begin{tabular}{l c c c c}
        \hline
        \textbf{Architecture} & \textbf{F1-score} & \textbf{Precision} & \textbf{Recall} & \textbf{Accuracy} \\
        \hline
        CE + Dice Loss & 0.9202 & 0.9183 & 0.9222 & 0.9759 \\
        Base CE & 0.9172 & 0.9140 & 0.9204 & 0.9755 \\
        CRF & 0.9137 & 0.9082 & 0.9194 & 0.9747 \\
        CRF + Focal Loss & 0.9118 & 0.9068 & 0.9169 & 0.9746 \\
        Dice Loss & 0.8959 & 0.8983 & 0.8935 & 0.9686 \\
        Focal Loss & 0.8502 & 0.8121 & 0.8922 & 0.9589 \\
        \hline
    \end{tabular}
    \caption{BERT: Final F1-scores Across All Loss Functions (epoch 5).}
    \label{tab:bert_summary}
\end{table}

\noindent \textbf{Overall Performance Ranking.} The six BERT variants exhibit a clear hierarchy of performance. CE + Dice Loss emerges as the absolute top performer with $F1 = 0.9202$, establishing itself as the optimal architecture. The gap between the best and worst variant is substantial: $0.9202 - 0.8502 = 0.0700$ (7.00 percentage points), proving the critical impact of objective function selection.

\noindent The top three variants (CE + Dice, Base CE, CRF) cluster tightly, all achieving $F1 \ge 0.9137$. This indicates that BERT's massive parameter count already captures profound representations, leaving only marginal room for improvement through structural constraints (CRF) or metric-specific optimization.

\noindent In contrast, Focal Loss ($F1 = 0.8502$) degrades performance severely ($-6.70\%$ relative to CE+Dice). This confirms that class-weighting and hard-mining are detrimental here. Pure Dice Loss performs moderately ($0.8959$), proving it requires the foundational gradient stability of CE to shine.

\noindent \textbf{Precision-Recall Trade-offs.} Examining the metrics reveals distinct optimization behaviors:
\begin{itemize}
    \item \textbf{CE + Dice Loss} achieves the ultimate balance ($P = 0.9183, R = 0.9222$), driving the highest overall F1.
    \item \textbf{Base CE} and \textbf{CRF} maintain highly comparable balances, slightly favoring recall over precision.
    \item \textbf{Dice Loss alone} flips the bias ($P = 0.8983, R = 0.8934$), being the only model where precision strictly exceeds recall.
    \item \textbf{Focal Loss} catastrophic precision drop ($P = 0.8121$) shows it over-emphasizes recall ($R = 0.8921$) due to over-penalization of the background class.
\end{itemize}

\noindent \textbf{Convergence and Stability.} All top variants demonstrate rapid monotonic improvement, effectively converging by epoch 3 or 4. CE + Dice shows a massive initial leap from epoch 1 to 2, indicating highly synergistic learning. Paradoxically, the worst-performing Focal Loss maintains the lowest absolute validation loss throughout training, highlighting a dangerous disconnect between raw log-loss minimization and actual discrete F1 metric maximization.

\noindent \textbf{Conclusion.} BERT-base fine-tuned with the hybrid \textbf{CE + Dice Loss} represents the absolute optimal configuration, achieving $F1 = 0.9202$. The tight cluster of top results underscores that the baseline CE is already highly effective. Complex sequence decoding (CRF) yields negligible empirical gains, and Focal Loss actively harms extraction precision. For dense Transformer architectures on properly formatted NER datasets, combining stable token-wise probabilities (CE) with global boundary optimization (Dice) yields the most robust, balanced, and state-of-the-art results.

\subsection{RoBERTa Results}

\subsubsection{Model Configuration.}
We train RoBERTa (\texttt{roberta-base}) for NER token classification.
RoBERTa-base is a strong Transformer encoder with 12 layers, 768 hidden dimensions, and approximately 125M parameters.
Compared with BERT, RoBERTa is trained with improved pre-training strategies and larger-scale data, making it highly competitive for token-level sequence labeling tasks.
Its byte-level BPE tokenizer is configured with appropriate word alignment to preserve the original token labels during preprocessing.

\subsubsection{Baseline: CE Loss.}

\begin{table}[H]
\centering
\caption{RoBERTa with Base Cross-Entropy Loss (5 epochs).}
\begin{tabular}{lcccc}
\hline
\textbf{Epoch} & \textbf{Train Loss} & \textbf{Val Loss} & \textbf{F1} & \textbf{Precision} / \textbf{Recall} / \textbf{Acc.} \\
\hline
1 & 0.1956 & 0.0817 & 0.9069 & 0.9001 / 0.9138 / 0.9731 \\
2 & 0.0723 & 0.0770 & 0.9113 & 0.9113 / 0.9113 / 0.9744 \\
3 & 0.0572 & 0.0736 & 0.9160 & 0.9104 / 0.9217 / 0.9758 \\
4 & 0.0492 & 0.0756 & 0.9197 & 0.9162 / 0.9233 / 0.9767 \\
5 & 0.0396 & 0.0773 & 0.9185 & 0.9157 / 0.9212 / 0.9769 \\
\hline
\end{tabular}
\label{tab:roberta_base_ce}
\end{table}

The baseline RoBERTa model achieves a strong final \textbf{F1-score of 0.9185}, already surpassing the corresponding DistilBERT baseline and slightly exceeding the BERT baseline.
Training loss decreases consistently over all five epochs, and validation loss remains low and stable, indicating efficient convergence.
Precision and recall are highly balanced ($P = 0.9157$, $R = 0.9212$), while token accuracy reaches \textbf{97.69\%}.
These results confirm that RoBERTa provides a very strong foundation for the NER task.
---

\subsubsection{Focal Loss.}

\begin{table}[H]
\centering
\caption{RoBERTa with Focal Loss ($\gamma = 2.0$, weighted by class frequency).}
\begin{tabular}{lcccc}
\hline
\textbf{Epoch} & \textbf{Train Loss} & \textbf{Val Loss} & \textbf{F1} & \textbf{Precision} / \textbf{Recall} / \textbf{Acc.} \\
\hline
1 & 0.1475 & 0.0427 & 0.8018 & 0.7427 / 0.8712 / 0.9379 \\
2 & 0.0358 & 0.0418 & 0.8257 & 0.7807 / 0.8763 / 0.9519 \\
3 & 0.0273 & 0.0398 & 0.8264 & 0.7846 / 0.8729 / 0.9492 \\
4 & 0.0205 & 0.0400 & 0.8454 & 0.8067 / 0.8879 / 0.9576 \\
5 & 0.0160 & 0.0419 & 0.8535 & 0.8176 / 0.8926 / 0.9600 \\
\hline
\end{tabular}
\label{tab:roberta_focal}
\end{table}

Focal Loss achieves a final \textbf{F1-score of 0.8535}, which is the lowest among all RoBERTa variants.
Although validation loss is numerically small, the model suffers a substantial drop in precision ($P = 0.8176$), while recall remains relatively high ($R = 0.8926$).
This imbalance suggests that Focal Loss over-emphasizes hard examples and produces too many false positive entity predictions.
Thus, as observed with BERT and DistilBERT, Focal Loss alone is not well-suited for this dataset.
---

\subsubsection{Dice Loss.}

\begin{table}[H]
\centering
\caption{RoBERTa with Dice Loss ($\epsilon = 10^{-6}$).}
\begin{tabular}{lcccc}
\hline
\textbf{Epoch} & \textbf{Train Loss} & \textbf{Val Loss} & \textbf{F1} & \textbf{Precision} / \textbf{Recall} / \textbf{Acc.} \\
\hline
1 & 0.1799 & 0.0921 & 0.8884 & 0.8860 / 0.8909 / 0.9664 \\
2 & 0.0912 & 0.0903 & 0.8972 & 0.9054 / 0.8893 / 0.9686 \\
3 & 0.0838 & 0.0854 & 0.9032 & 0.9037 / 0.9027 / 0.9707 \\
4 & 0.0756 & 0.0843 & 0.9084 & 0.9125 / 0.9043 / 0.9717 \\
5 & 0.0761 & 0.0816 & 0.9090 & 0.9125 / 0.9055 / 0.9725 \\
\hline
\end{tabular}
\label{tab:roberta_dice}
\end{table}

Dice Loss achieves a final \textbf{F1-score of 0.9090}, showing solid and stable performance throughout training.
The model converges smoothly and maintains a well-balanced precision-recall profile ($P = 0.9125$, $R = 0.9055$).
Compared with the CE baseline, however, standalone Dice Loss performs slightly worse, suggesting that F1-oriented optimization alone is not sufficient to outperform standard cross-entropy on this task.
---

\subsubsection{CE + Dice Loss.}

\begin{table}[H]
\centering
\caption{RoBERTa with Combined CE + Dice Loss.}
\begin{tabular}{lcccc}
\hline
\textbf{Epoch} & \textbf{Train Loss} & \textbf{Val Loss} & \textbf{F1} & \textbf{Precision} / \textbf{Recall} / \textbf{Acc.} \\
\hline
1 & 0.3989 & 0.1784 & 0.9091 & 0.9023 / 0.9161 / 0.9745 \\
2 & 0.1683 & 0.1688 & 0.9166 & 0.9182 / 0.9151 / 0.9755 \\
3 & 0.1422 & 0.1633 & 0.9192 & 0.9197 / 0.9187 / 0.9765 \\
4 & 0.1238 & 0.1640 & 0.9218 & 0.9196 / 0.9241 / 0.9768 \\
5 & 0.1103 & 0.1631 & 0.9223 & 0.9215 / 0.9231 / 0.9772 \\
\hline
\end{tabular}
\label{tab:roberta_ce_dice}
\end{table}

The combined CE + Dice Loss achieves the \textbf{best overall performance} among all RoBERTa variants, with a final \textbf{F1-score of 0.9223}.
This model also achieves the highest precision ($P = 0.9215$) while maintaining very strong recall ($R = 0.9231$).
The combination of CE and Dice successfully integrates stable token-level optimization with F1-aware structural matching, resulting in the most balanced and robust model.
---

\subsubsection{CRF Layer.}

\begin{table}[H]
\centering
\caption{RoBERTa + CRF (no focal loss).}
\begin{tabular}{lcccc}
\hline
\textbf{Epoch} & \textbf{Train Loss} & \textbf{Val Loss} & \textbf{F1} & \textbf{Precision} / \textbf{Recall} / \textbf{Acc.} \\
\hline
1 & 5.0027 & 1.8676 & 0.9027 & 0.8923 / 0.9135 / 0.9716 \\
2 & 1.6522 & 1.7130 & 0.9116 & 0.9087 / 0.9145 / 0.9742 \\
3 & 1.2982 & 1.6659 & 0.9147 & 0.9064 / 0.9231 / 0.9754 \\
4 & 1.1609 & 1.6694 & 0.9201 & 0.9150 / 0.9253 / 0.9763 \\
5 & 0.9152 & 1.6881 & 0.9220 & 0.9195 / 0.9246 / 0.9772 \\
\hline
\end{tabular}
\label{tab:roberta_crf}
\end{table}

The RoBERTa + CRF model performs extremely well, achieving a final \textbf{F1-score of 0.9220}, which is only slightly below the CE + Dice variant.
It records the highest recall ($R = 0.9246$) and the highest token accuracy (\textbf{97.72\%}) among all RoBERTa models.
This demonstrates the strength of CRF-based sequence modeling, especially in enforcing valid label transitions and improving entity consistency.
---

\subsubsection{CRF + Focal Loss.}

\begin{table}[H]
\centering
\caption{RoBERTa + CRF with Focal Loss ($\gamma = 2.0$).}
\begin{tabular}{lcccc}
\hline
\textbf{Epoch} & \textbf{Train Loss} & \textbf{Val Loss} & \textbf{F1} & \textbf{Precision} / \textbf{Recall} / \textbf{Acc.} \\
\hline
1 & 4.7636 & 1.6471 & 0.9058 & 0.8949 / 0.9170 / 0.9722 \\
2 & 1.4418 & 1.4699 & 0.9103 & 0.9062 / 0.9145 / 0.9744 \\
3 & 1.1087 & 1.4355 & 0.9137 & 0.9063 / 0.9212 / 0.9751 \\
4 & 0.9837 & 1.4112 & 0.9186 & 0.9123 / 0.9249 / 0.9765 \\
5 & 0.7423 & 1.4216 & 0.9203 & 0.9172 / 0.9234 / 0.9769 \\
\hline
\end{tabular}
\label{tab:roberta_crf_focal}
\end{table}

The CRF + Focal Loss variant reaches a final \textbf{F1-score of 0.9203}, which is competitive but still below both CE + Dice and pure CRF.
Its precision and recall remain strong ($P = 0.9172$, $R = 0.9234$), but the addition of Focal Loss does not produce further gains.
This suggests that, as in the other architectures, Focal Loss introduces optimization pressure that is not fully compatible with the CRF objective.
---

\subsubsection{Summary: All RoBERTa Variants.}

\begin{table}[htbp]
\centering
\caption{RoBERTa: Final F1-scores Across All Loss Functions (epoch 5).}
\begin{tabular}{lcccc}
\hline
\textbf{Architecture} & \textbf{F1-score} & \textbf{Precision} & \textbf{Recall} & \textbf{Accuracy} \\
\hline
CE + Dice Loss & \textbf{0.9223} & \textbf{0.9215} & 0.9231 & 0.9772 \\
CRF & 0.9220 & 0.9195 & \textbf{0.9246} & \textbf{0.9772} \\
CRF + Focal Loss & 0.9203 & 0.9172 & 0.9234 & 0.9769 \\
Base CE & 0.9185 & 0.9157 & 0.9212 & 0.9769 \\
Dice Loss & 0.9090 & 0.9125 & 0.9055 & 0.9725 \\
Focal Loss & 0.8535 & 0.8176 & 0.8926 & 0.9600 \\
\hline
\end{tabular}
\label{tab:roberta_summary}
\end{table}

\paragraph{Overall Performance Ranking.}
The six RoBERTa variants show a clear ranking pattern, with \textbf{CE + Dice Loss emerging as the best-performing configuration} at $F1 = 0.9223$.
The performance gap between the best and worst variant is substantial: $0.9223 - 0.8535 = 0.0688$ (6.88 percentage points), confirming again that loss function selection strongly affects the final extraction quality.

The top four variants (CE + Dice, CRF, CRF + Focal Loss, and Base CE) form a very tight performance cluster, all achieving $F1 \geq 0.9185$.
This indicates that RoBERTa provides highly expressive contextual representations, leaving only limited room for improvement through additional sequence modeling or specialized objectives.
Among them, CE + Dice achieves the best balance, while CRF reaches the best recall and accuracy.

By contrast, \textbf{Focal Loss} performs much worse than all other RoBERTa variants.
Its final F1-score of 0.8535 is nearly 7 points lower than the best configuration, demonstrating that hard-example emphasis is again detrimental rather than helpful for this NER setting.
Standalone Dice Loss performs reasonably well ($F1 = 0.9090$), but it also remains below the stronger CE-based hybrids.

\paragraph{Precision-Recall Trade-offs.}
The precision-recall statistics reveal meaningful behavioral differences:

\begin{itemize}
\item \textbf{CE + Dice Loss} achieves the best overall precision-recall balance ($P = 0.9215$, $R = 0.9231$), producing the highest F1-score.
\item \textbf{CRF} slightly favors recall ($R = 0.9246$), which helps it match the best token accuracy and nearly match the best F1-score.
\item \textbf{Base CE} maintains a stable and balanced profile, confirming that standard cross-entropy is already highly effective with RoBERTa.
\item \textbf{Dice Loss alone} slightly favors precision over recall ($P = 0.9125$, $R = 0.9055$), suggesting more conservative entity predictions.
\item \textbf{Focal Loss} again shows a severe precision drop ($P = 0.8176$) while recall remains relatively high, indicating over-prediction of positive labels and excessive false positives.
\end{itemize}

\paragraph{Convergence and Stability.}
All strong RoBERTa variants converge quickly and smoothly, with most performance gains achieved by epoch 3 or 4.
\textbf{CE + Dice Loss} shows consistent improvement across all epochs and reaches the highest final F1-score.
\textbf{CRF} also converges stably, though its validation loss remains numerically larger due to the different scale of CRF negative log-likelihood; therefore, direct loss comparisons between CRF and non-CRF models should be interpreted with caution.

As observed in the BERT and DistilBERT experiments, \textbf{Focal Loss} maintains relatively low numerical validation loss but yields the weakest entity-level F1-score.
This again highlights that minimizing raw loss does not necessarily correspond to maximizing the seqeval metric.

\paragraph{Conclusion.}
RoBERTa fine-tuned with \textbf{CE + Dice Loss} is the strongest RoBERTa configuration, achieving $F1 = 0.9223$, precision $= 0.9215$, recall $= 0.9231$, and token accuracy $= 97.72\%$.
RoBERTa + CRF is an extremely close second, with the best recall and matched best accuracy.
These findings suggest that for RoBERTa, both hybrid token-level optimization (CE + Dice) and explicit sequence decoding (CRF) are highly effective, but CE + Dice provides the most balanced and reliable overall solution.
As with the other backbones, Focal Loss alone remains the weakest choice.

\subsection{DistilBERT results}

\subsubsection{Model Configuration.}
We train DistilBERT (\texttt{distilbert-base-cased}) for NER token classification.
DistilBERT is a lightweight distilled variant of BERT with 6 layers, 768 hidden dimensions, and 66M parameters,
making it computationally efficient while retaining strong performance. The tokenizer is configured without
prefix spacing (unlike RoBERTa).

\subsubsection{Baseline: CE Loss.}

\begin{table}[H]
\centering
\caption{DistilBERT with Base Cross-Entropy Loss (5 epochs).}
\begin{tabular}{lcccc}
\hline
\textbf{Epoch} & \textbf{Train Loss} & \textbf{Val Loss} & \textbf{F1} & \textbf{Precision} / \textbf{Recall} / \textbf{Acc.} \\
\hline
1 & 0.2308 & 0.0940 & 0.8929 & 0.8809 / 0.9053 / 0.9697 \\
2 & 0.0763 & 0.0844 & 0.9027 & 0.9004 / 0.9050 / 0.9722 \\
3 & 0.0574 & 0.0817 & 0.9069 & 0.9021 / 0.9117 / 0.9737 \\
4 & 0.0480 & 0.0840 & 0.9104 & 0.9049 / 0.9160 / 0.9741 \\
5 & 0.0377 & 0.0859 & 0.9095 & 0.9046 / 0.9145 / 0.9744 \\
\hline
\end{tabular}
\label{tab:distilbert_base_ce}
\end{table}

The baseline cross-entropy model achieves a final \textbf{F1-score of 0.9095}, providing a strong reference for comparison. 
Training loss decreases steadily from 0.23 to 0.04 across five epochs, while validation loss remains stable (0.084--0.094), indicating good convergence with minimal overfitting. 
Precision and recall are well-balanced ($P \approx 0.905$, $R \approx 0.914$), and token accuracy reaches \textbf{97.44\%}. 
Overall, CE loss provides a stable and reliable baseline for the NER task.
---

\subsubsection{Focal Loss.}

\begin{table}[H]
\centering
\caption{DistilBERT with Focal Loss ($\gamma = 2.0$, weighted by class frequency).}
\begin{tabular}{lcccc}
\hline
\textbf{Epoch} & \textbf{Train Loss} & \textbf{Val Loss} & \textbf{F1} & \textbf{Precision} / \textbf{Recall} / \textbf{Acc.} \\
\hline
1 & 0.1784 & 0.0526 & 0.7528 & 0.6821 / 0.8399 / 0.9194 \\
2 & 0.0388 & 0.0435 & 0.8012 & 0.7504 / 0.8594 / 0.9440 \\
3 & 0.0263 & 0.0456 & 0.8265 & 0.7827 / 0.8755 / 0.9494 \\
4 & 0.0194 & 0.0438 & 0.8389 & 0.7956 / 0.8871 / 0.9546 \\
5 & 0.0141 & 0.0462 & 0.8447 & 0.8063 / 0.8869 / 0.9572 \\
\hline
\end{tabular}
\label{tab:distilbert_focal}
\end{table}

Focal Loss achieves a final \textbf{F1-score of 0.8447}, which is significantly lower than the CE baseline. 
Although training loss decreases smoothly, the model shows reduced precision (0.8063) and only moderate recall (0.8869). 
This suggests that emphasizing hard examples through Focal Loss may be unnecessary for this dataset and may reduce overall discriminative performance. 
Overall, Focal Loss underperforms compared to the baseline.
---

\subsubsection{Dice Loss.}

\begin{table}[H]
\centering
\caption{DistilBERT with Dice Loss ($\epsilon = 10^{-6}$).}
\begin{tabular}{lcccc}
\hline
\textbf{Epoch} & \textbf{Train Loss} & \textbf{Val Loss} & \textbf{F1} & \textbf{Precision} / \textbf{Recall} / \textbf{Acc.} \\
\hline
1 & 0.2118 & 0.1172 & 0.8599 & 0.8489 / 0.8712 / 0.9574 \\
2 & 0.1062 & 0.1028 & 0.8800 & 0.8839 / 0.8761 / 0.9654 \\
3 & 0.0918 & 0.0987 & 0.8883 & 0.8936 / 0.8831 / 0.9665 \\
4 & 0.0858 & 0.0964 & 0.8882 & 0.8887 / 0.8877 / 0.9679 \\
5 & 0.0822 & 0.0949 & 0.8911 & 0.8937 / 0.8884 / 0.9685 \\
\hline
\end{tabular}
\label{tab:distilbert_dice}
\end{table}

Dice Loss achieves a final \textbf{F1-score of 0.8911}, slightly below the CE baseline. 
The model converges quickly and maintains balanced precision and recall ($P = 0.8937$, $R = 0.8884$). 
Since Dice Loss directly optimizes an F1-like objective, it produces stable and balanced predictions, though it does not surpass the baseline CE performance.
---

\subsubsection{CE + Dice Loss.}

\begin{table}[H]
\centering
\caption{DistilBERT with Combined CE + Dice Loss.}
\begin{tabular}{lcccc}
\hline
\textbf{Epoch} & \textbf{Train Loss} & \textbf{Val Loss} & \textbf{F1} & \textbf{Precision} / \textbf{Recall} / \textbf{Acc.} \\
\hline
1 & 0.4448 & 0.2078 & 0.8880 & 0.8720 / 0.9046 / 0.9689 \\
2 & 0.1748 & 0.1844 & 0.9050 & 0.9023 / 0.9078 / 0.9724 \\
3 & 0.1411 & 0.1805 & 0.9069 & 0.9039 / 0.9100 / 0.9737 \\
4 & 0.1207 & 0.1798 & 0.9119 & 0.9076 / 0.9163 / 0.9744 \\
5 & 0.1051 & 0.1805 & 0.9109 & 0.9066 / 0.9152 / 0.9746 \\
\hline
\end{tabular}
\label{tab:distilbert_ce_dice}
\end{table}

The combined CE + Dice Loss achieves the \textbf{best performance} with a final \textbf{F1-score of 0.9109}. 
The model converges rapidly and maintains excellent precision-recall balance ($P = 0.9066$, $R = 0.9152$). 
This combination benefits from CE's stable probability learning and Dice's F1-oriented optimization, resulting in the most robust performance among all variants.
---

\subsubsection{CRF Layer.}

\begin{table}[H]
\centering
\caption{DistilBERT + CRF (no focal loss).}
\begin{tabular}{lcccc}
\hline
\textbf{Epoch} & \textbf{Train Loss} & \textbf{Val Loss} & \textbf{F1} & \textbf{Precision} / \textbf{Recall} / \textbf{Acc.} \\
\hline
1 & 6.2858 & 2.1382 & 0.8846 & 0.8713 / 0.8985 / 0.9677 \\
2 & 1.8468 & 1.9222 & 0.8971 & 0.8915 / 0.9027 / 0.9713 \\
3 & 1.3623 & 1.8682 & 0.8990 & 0.8898 / 0.9084 / 0.9717 \\
4 & 1.1910 & 1.8535 & 0.9046 & 0.8957 / 0.9136 / 0.9735 \\
5 & 0.9343 & 1.8865 & 0.9064 & 0.9003 / 0.9126 / 0.9738 \\
\hline
\end{tabular}
\label{tab:distilbert_crf}
\end{table}

The DistilBERT + CRF model achieves a final \textbf{F1-score of 0.9064}, slightly below the CE baseline. 
CRF models label dependencies at the sequence level, which ensures valid tag transitions and balanced predictions ($P = 0.9003$, $R = 0.9126$). 
However, the improvement over the baseline is limited, suggesting that the base model already captures most sequence structure in this dataset.
---

\subsubsection{CRF + Focal Loss.}

\begin{table}[H]
\centering
\caption{DistilBERT + CRF with Focal Loss ($\gamma = 2.0$).}
\begin{tabular}{lcccc}
\hline
\textbf{Epoch} & \textbf{Train Loss} & \textbf{Val Loss} & \textbf{F1} & \textbf{Precision} / \textbf{Recall} / \textbf{Acc.} \\
\hline
1 & 6.0633 & 1.8933 & 0.8858 & 0.8724 / 0.8996 / 0.9679 \\
2 & 1.6268 & 1.6997 & 0.8959 & 0.8898 / 0.9021 / 0.9707 \\
3 & 1.1657 & 1.6224 & 0.8979 & 0.8896 / 0.9063 / 0.9717 \\
4 & 1.0049 & 1.5945 & 0.9030 & 0.8940 / 0.9122 / 0.9733 \\
5 & 0.7580 & 1.6152 & 0.9049 & 0.8982 / 0.9117 / 0.9736 \\
\hline
\end{tabular}
\label{tab:distilbert_crf_focal}
\end{table}

The CRF + Focal Loss variant achieves a final \textbf{F1-score of 0.9049}, slightly lower than the CRF-only model. 
Although training loss decreases steadily, the combination of sequence-level CRF optimization and token-level Focal Loss does not improve performance. 
The results suggest that the two objectives may introduce conflicting optimization signals, making this combination less effective for the task.

\subsubsection{Summary: All DistilBERT Variants.}

\begin{table}[htbp]
\centering
\caption{DistilBERT: Final F1-scores Across All Loss Functions (epoch 5).}
\begin{tabular}{lcccc}
\hline
\textbf{Architecture} & \textbf{F1-score} & \textbf{Precision} & \textbf{Recall} & \textbf{Accuracy} \\
\hline
CE + Dice Loss & \textbf{0.9109} & \textbf{0.9066} & \textbf{0.9152} & \textbf{0.9746} \\
Base CE & 0.9095 & 0.9046 & 0.9145 & 0.9744 \\
CRF & 0.9064 & 0.9003 & 0.9126 & 0.9738 \\
CRF + Focal Loss & 0.9049 & 0.8982 & 0.9117 & 0.9736 \\
Dice Loss & 0.8911 & 0.8937 & 0.8884 & 0.9685 \\
Focal Loss & 0.8447 & 0.8063 & 0.8869 & 0.9572 \\
\hline
\end{tabular}
\label{tab:distilbert_summary}
\end{table}

\paragraph{Overall Performance Ranking.}
The six DistilBERT variants exhibit a clear hierarchy of performance.
\textbf{CE + Dice Loss emerges as the top performer} with $F1 = 0.9109$, establishing itself as the optimal choice for this NER task.
The performance gap between the best and worst variant is substantial: $0.9109 - 0.8447 = 0.0662$ (6.62 percentage points), highlighting the significant impact of loss function design on final performance.

The top three variants (CE + Dice, Base CE, CRF) cluster tightly within 0.0045 F1 points, with all achieving $F1 \geq 0.9064$.
This suggests that for DistilBERT on this dataset, sophisticated sequence modeling (CRF) and F1-aware losses (Dice) provide only marginal improvements over the simple baseline—a testament to the strong baseline and well-structured nature of the NER task.

In contrast, Focal Loss ($F1 = 0.8447$) and Dice Loss alone ($F1 = 0.8911$) show significantly degraded performance.
Focal Loss's poor performance ($-6.48\%$ relative to CE + Dice) is particularly noteworthy, as it suggests that class weighting and hard example emphasis are either unnecessary or harmful on a balanced dataset.
Dice Loss performs moderately ($-1.98\%$ relative to CE + Dice), indicating that while F1-aware optimization is beneficial in isolation, it achieves superior results when combined with cross-entropy rather than used alone.

\paragraph{Precision-Recall Trade-offs.}
Examining the precision-recall curves reveals important trade-offs:

\begin{itemize}
\item \textbf{CE + Dice Loss} achieves the best balance ($P = 0.9066$, $R = 0.9152$), with recall slightly exceeding precision, resulting in the highest $F1$.
\item \textbf{Base CE} and \textbf{CRF} maintain comparable balances ($P \approx 0.9025$, $R \approx 0.9136$), differing only marginally from CE + Dice.
\item \textbf{Dice Loss alone} exhibits reversed bias ($P = 0.8937$, $R = 0.8884$), where precision exceeds recall, suggesting the Dice objective favors entity detection over boundary accuracy.
\item \textbf{Focal Loss} severely degrades precision ($P = 0.8063$), indicating it over-emphasizes recall ($R = 0.8869$) at the expense of precision, likely due to aggressive down-weighting of easy examples.
\item \textbf{CRF + Focal Loss} falls between CRF and Focal Loss in precision ($P = 0.8982$), inheriting some precision loss from Focal Loss despite CRF's structural constraints.
\end{itemize}

\paragraph{Convergence and Stability.}
All variants exhibit smooth, monotonic improvement in $F1$ across epochs, with convergence achieved by epoch 3--4.
\textbf{CE + Dice Loss} demonstrates the most dramatic jump in $F1$ from epoch 1 ($0.8880$) to epoch 2 ($0.9050$), indicating efficient learning from the combined loss signals.
\textbf{Focal Loss} and \textbf{Dice Loss} show slower convergence and lower epoch-wise $F1$ values throughout training, suggesting these objectives require more careful tuning or are fundamentally misaligned with the task structure.

Validation loss patterns confirm stability: CE + Dice, Base CE, CRF, and CRF + Focal Loss all stabilize after epoch 2 with minimal fluctuation, whereas Focal Loss maintains the lowest absolute validation loss but yields the worst $F1$, indicating a disconnect between loss minimization and metric maximization.

\paragraph{Conclusion.}
DistilBERT with CE + Dice Loss represents the optimal configuration among the tested variants, achieving state-of-the-art performance on this NER task with $F1 = 0.9109$, precision $= 0.9066$, recall $= 0.9152$, and token accuracy $= 97.46\%$.
The narrow performance gap between the top variants (CE + Dice, Base CE, CRF) underscores that the baseline is already highly effective, leaving limited room for improvement via advanced techniques.
More sophisticated losses and decoding strategies contribute only marginal gains, and in the case of Focal Loss, degrade performance.
These findings suggest that the NER task at hand exhibits characteristics favorable to standard approaches: relatively balanced class distribution, clean label sequences, and sufficient model capacity in DistilBERT to capture linguistic patterns without special handling of imbalance or sequence constraints.


\subsection{Model Comparison and Discussion}

\subsubsection{Cross-Architecture Performance Comparison}

\begin{table}[H]
    \centering
    \caption{All Model Variants Across All Architectures (Final Epoch Results).}
    \small
    \begin{tabular}{l l c c c c c}
        \hline
        \textbf{Architecture} & \textbf{Loss Function} & \textbf{Params} & \textbf{F1-score} & \textbf{Precision} & \textbf{Recall} & \textbf{Accuracy} \\
        \hline
        \multicolumn{7}{c}{\textbf{BERT-base (110M Parameters)}} \\
        \hline
        BERT-base & CE + Dice Loss & 110M & \textbf{0.9202} & 0.9183 & 0.9222 & 0.9759 \\
        BERT-base & Base CE & 110M & 0.9172 & 0.9140 & 0.9204 & 0.9755 \\
        BERT-base & CRF & 110M & 0.9137 & 0.9082 & 0.9194 & 0.9747 \\
        BERT-base & CRF + Focal Loss & 110M & 0.9118 & 0.9068 & 0.9169 & 0.9746 \\
        BERT-base & Dice Loss & 110M & 0.8959 & 0.8983 & 0.8935 & 0.9686 \\
        BERT-base & Focal Loss & 110M & 0.8502 & 0.8121 & 0.8922 & 0.9589 \\
        \hline
        \multicolumn{7}{c}{\textbf{RoBERTa-base (125M Parameters)}} \\
        \hline
        RoBERTa-base & CE + Dice Loss & 125M & \textbf{0.9223} & \textbf{0.9215} & 0.9231 & 0.9772 \\
        RoBERTa-base & CRF & 125M & 0.9220 & 0.9195 & \textbf{0.9246} & \textbf{0.9772} \\
        RoBERTa-base & CRF + Focal Loss & 125M & 0.9203 & 0.9172 & 0.9234 & 0.9769 \\
        RoBERTa-base & Base CE & 125M & 0.9185 & 0.9157 & 0.9212 & 0.9769 \\
        RoBERTa-base & Dice Loss & 125M & 0.9090 & 0.9125 & 0.9055 & 0.9725 \\
        RoBERTa-base & Focal Loss & 125M & 0.8535 & 0.8176 & 0.8926 & 0.9600 \\
        \hline
        \multicolumn{7}{c}{\textbf{DistilBERT-base (66M Parameters)}} \\
        \hline
        DistilBERT-base & CE + Dice Loss & 66M & \textbf{0.9109} & \textbf{0.9066} & \textbf{0.9152} & \textbf{0.9746} \\
        DistilBERT-base & Base CE & 66M & 0.9095 & 0.9046 & 0.9145 & 0.9744 \\
        DistilBERT-base & CRF & 66M & 0.9064 & 0.9003 & 0.9126 & 0.9738 \\
        DistilBERT-base & CRF + Focal Loss & 66M & 0.9049 & 0.8982 & 0.9117 & 0.9736 \\
        DistilBERT-base & Dice Loss & 66M & 0.8911 & 0.8937 & 0.8884 & 0.9685 \\
        DistilBERT-base & Focal Loss & 66M & 0.8447 & 0.8063 & 0.8869 & 0.9572 \\
        \hline
    \end{tabular}
    \label{tab:comprehensive_all_models}
\end{table}

\paragraph{Key Findings}

\textbf{Best Configuration:} RoBERTa-base + CE + Dice Loss achieves the highest F1-score (0.9223) with balanced precision and recall.

\textbf{Performance Hierarchy:} RoBERTa (125M) $\to$ BERT (110M, -0.0021 F1) $\to$ DistilBERT (66M, -0.0093 F1). Despite 47\% fewer parameters, DistilBERT remains competitive, making it ideal for resource-constrained deployment.

\textbf{Loss Function Consistency:} CE + Dice Loss outperforms alternatives across all architectures, while Focal Loss consistently underperforms (-6--7\% F1), indicating the findings are architecture-agnostic.

\textbf{Precision-Recall Balance:} CE + Dice achieves the most balanced profiles, with a slight recall bias (0.004--0.0049 F1 points higher than precision) beneficial for NER applications.

\paragraph{Recommendations}

\begin{itemize}
    \item \textbf{Maximum Accuracy:} RoBERTa-base + CE + Dice (F1 = 0.9223)
    \item \textbf{Production (Best Trade-off):} BERT-base + CE + Dice (F1 = 0.9202, 12\% fewer parameters)
    \item \textbf{Edge/Latency-Critical:} DistilBERT-base + CE + Dice (F1 = 0.9109, 47\% parameter reduction)
    \item \textbf{Loss Function:} Always use CE + Dice Loss; avoid Focal Loss for balanced NER datasets
\end{itemize}

\paragraph{Conclusion}

RoBERTa + CE + Dice Loss is the best-performing configuration (F1 = 0.9223). However, minimal performance gaps among top variants suggest that architecture choice should prioritize deployment constraints (latency, memory, cost) over marginal accuracy differences.